Fractional Types

### 2013-02-23 frac.tex
_

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{compact closed category} from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle.  Our main novel
technical device to achieving the higher-order extension is a
\emph{fractional type}, which we so named because of its duality with
conventional product types.  By adding a formal (linear) dual to (type
theoretic) products, we can internalize the notion that ``taking an input''
corresponds to negative information, which turns out to be the basis of
first-class functions. The remainder of the paper reviews {{Pi}} and then
introduces the syntax and semantics of the extension with fractional
types. We then study the properties of the extended language and establish
its expressiveness via several constructions and examples that exploit its
ability to model higher-order computations.

@@

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b 
\noindent Types include the unit type {{1}}, sum types {{b1 + b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary: we do not treat them as first-class citizens,
and they only occur when observing the effect of an isomorphism.

@@

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
however some informal intuition about the \emph{meaning} of such fractional
types and values. (See Sec.~\ref{sec:cat} for more discussion about this
point.) In more detail, given a symmetric monoidal category modeling a
certain language (like {{Pi}}), one can obtain a higher-order extension of
the language by extending the category to a \emph{compact closed} one, i.e.,
a category in which morphisms are representable as objects. The key to this
extension is to add dual objects (in our case the fractionals) satisfying the
following isomorphism witnessed by two new type indexed combinators 
{{eta*_b}} and {{eps*_b}}:
%subcode{bnf} include main
%! columnStyle = r r c l l
%eta*_b :&  1 & <-> & 1/b * b &: eps*_b
%% XXX: diagrams need unit wires(?)
\begin{multicols}{2}
  \dgm{eta_times2}

@@

\noindent From a programming perspective, we think of a value {{1/v}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{v}} or as a \emph{pattern} representing the absence of
some information of a specific shape (i.e., \emph{negative information}),
that can only be reconciliated by an actual value {{v}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
out of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information.

>>>>

Technically speaking, obtaining a higher-order extension requires the
construction of a 
%\emph{compact closed category}
\emph{closed category} 
%
from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle.  Our main novel
technical device to achieving the higher-order extension is a
\emph{fractional type}, which we so named because of its duality with
conventional product types.  By adding a formal (linear) dual to (type
theoretic) products, we can internalize the notion that ``taking an input''
corresponds to negative information, which turns out to be the basis of
first-class functions. \roshan{The last sentence is terrible.} 
The remainder of the paper reviews {{Pi}} and then
introduces the syntax and semantics of the extension with fractional
types. We then study the properties of the extended language and establish
its expressiveness via several constructions and examples that exploit its
ability to model higher-order computations.

@@

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b 
\noindent 
Types include the unit type {{1}}, sum types {{b1+b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary: we do not treat them as first-class citizens,
and they only occur when observing the effect of an isomorphism.

@@

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
however some informal intuition about the \emph{meaning} of such fractional
types and values. (See Sec.~\ref{sec:cat} for more discussion about this
point.) In more detail, given a symmetric monoidal category modeling a
certain language (like {{Pi}}), one can obtain a higher-order extension of
the language by extending the category to a \emph{compact closed} one, i.e.,
a category in which morphisms are representable as objects. The key to this
extension is to add dual objects (in our case the fractionals) satisfying the
following isomorphism witnessed by two new type indexed combinators 
{{eta*_b}} and {{eps*_b}}:
\begin{multicols}{3}
  \dgm{eta_times2}

\columnbreak

%subcode{bnf} include main
%! columnStyle = r r c l l
%eta*_b :&  1 & <-> & 1/b * b &: eps*_b
%% XXX: diagrams need unit wires(?)
%% Roshan: We should just be clear that unit wires are dropped. Dgms
%% including them dont communicate any additional information and
%% appear crowded.

\columnbreak

@@

\vspace{-10pt}
\noindent From a programming perspective, we think of a value {{1/v}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{v}} or as a \emph{pattern} representing the absence of
some information of a specific shape (i.e., \emph{negative information}),
that can only be reconciliated by an actual value {{v}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
out of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information.
